\textsc{\wobble: a data-driven method for precision radial velocities}

### 2018-09-10 paper/paper.tex
writing

The model underpinning \wobble is designed to be flexible and easily extensible to a variety of use cases. However, a few rigid assumptions are necessary, and we outline those here.

We take data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\log f$ for pixel $m$ of M at epoch $n$ of N. We also have a corresponding $M \times N$ matrix of wavelength solutions obtained in the course of instrument calibrations, which we assume to be accurate. Again, the entries of this matrix are actually in logarithmic form ($\log \lambda$). This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.

@@

Given these assumptions, we can model each data column $y_n$ as:
$$y_n = P(v_{\star, n}) \mu_{\star} + \mu_{t} + W_{t} z_n + noise,$$
where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu$ is the mean spectral template for each component.  In addition to its mean spectrum $\mu_t$, the telluric component also takes a time-dependent component assembled from two variables: $W$, a matrix of ``basis vectors,'' is weighted by $z_n$ to form the spectral contribution at epoch $n$. 
Each $\mu$ has length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid. 
Correspondingly, $W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications.

Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
$$ \log \lambda(v) = \log \lambda_{0} + \log \left(\frac{1 - v/c}{1 + v/c}\right).
$$
This shift is applied to the $M'$-length wavelength grid for the template spectrum before interpolating the template to the $M$-length data space.

%With these models in hand, we are then able to infer an \RV\ shift for both model components at each epoch.
We evaluate a likelihood for each epoch
$$ \ln \mathcal{L_n} = \sum_{m} -0.5 (y_{m,n} - y_{m',n})^T C^{-1} (y_{m,n}-y_{m',n}),
$$
comparing our model $y'$ to the data $y$, with $C$ representing our covariance matrix of the uncertainties on each data point.
Our model is agnostic to the intrinsic wavelength of any individual stellar or telluric feature, so while we infer the  Doppler shift relative to our model, it does not apply any information about the absolute \RV of the star relative to the solar system. 
To mitigate the potential degeneracy of developing a model effectively redshifted relative to all observations, we add (in log space) a Gaussian prior to our likelihood value on the mean \RV for all $N$ epochs such that
$$ P(m|X) = \ln \mathcal{L}  -0.5 * \frac{\sum(v)^2}{(Nl)^2}.
$$
\todo{someone should define these variables}

We build an initial template for the stellar spectrum by setting $v_{\star} = 0$ for all epochs. We can then Doppler shift the spectra by the barycentric velocity \BERV and calculate the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. The telluric template can then be initialized using the residuals after the stellar contribution has been removed, again by binning (this time without applying any Doppler shift to the spectra) and taking the median values of each bin. Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed. The $K$ highest eigenweights and their corresponding eigenvectors were taken as the basis weights and vectors. \todo{(math here?)}

@@

The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, using the \TF package \todo{[terminology?]} in \code{python}. \TF is a \todo{[blah blah technical blah]}. While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this use case as well. In particular, the necessary optimizations can be performed with high efficiency by \TF: the below-described analysis of \todo{x} spectra of 51 Peg runs in \todo{y} hours on a standard Macbook laptop.

>>>>

The model underpinning \wobble is designed to be flexible and easily extensible to a variety of situations. 
However, a few rigid assumptions are necessary, and we outline those here.

We take data to be the $M \times N$ matrix Y, where each entry $y_{m,n}$ is the observed logarithmic flux $\ln f$ for pixel $m$ of $M$ at epoch $n$ of $N$. 
We also have a corresponding $M \times N$ matrix of wavelength solutions obtained in the course of instrument calibrations, which we assume to be accurate.
Again, the entries of this matrix are in logarithmic form ($\ln \lambda$). 
This choice makes the Doppler shift an additive term, considerably simplifying the math to follow.

@@

Given these assumptions, we can model each data column $y_n$ as the sum of the stellar and telluric spectra:
\begin{equation}
y_n = y_{\star, n} + y_{t, n} + noise.
\end{equation}

The stellar spectrum contribution is: 
\begin{equation}
y_{\star, n} = P(v_{obs, n}) \mu_{\star},
\end{equation}
where $P$ is an interpolation operator which applies a Doppler shift by velocity $v_n$ and $\mu_{\star}$ is a spectral template with length $M'$ corresponding to the length of an arbitrarily-chosen wavelength grid. 
Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
\begin{equation}
 \ln \lambda(v) = \ln \lambda_{0} + \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}
The $P$ operator applies this term to the $M'$-length wavelength grid for the template spectrum before interpolating the template to the $M$-length data space.

The telluric spectrum contribution is:
\begin{equation}
y_{t, n} =  A_n(\mu_{t} + W_{t} z_n)
\end{equation}
In addition to its mean spectrum $\mu_t$, the telluric component also takes a time-dependent component assembled from two variables: $W$, a matrix of ``basis vectors,'' is weighted by $z_n$ to form the spectral contribution at epoch $n$. 
$W$ has the shape $M' \times K$ and $z_n$ is a $K$-vector, where $K$ is the number of basis vectors used. 
For the purposes of this work we found good performance with $K$ set to 3, but this may vary for other applications. 
The net telluric spectrum (mean + time-variable components) is weighted by the air mass at the time of observation $A_n$, a known quantity.

We evaluate a likelihood for each epoch
\begin{equation}
\ln \mathcal{L}_n = \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C^{-1} (y_{m,n}-y'_{m,n}),
\end{equation}
comparing our model $y'$ to the data $y$, with $C$ representing the covariance matrix of uncertainties on each data point.
Our model is agnostic to the intrinsic wavelength of any individual stellar or telluric feature, so while we infer the Doppler shift relative to our model, it does not apply any information about the absolute \RV of the star relative to the solar system. 
%To mitigate the potential degeneracy of developing a model effectively redshifted relative to all observations, we add (in log space) a Gaussian prior to our likelihood value on the mean \RV for all $N$ epochs such that
%$$ P(m|X) = \ln \mathcal{L}  -0.5 * \frac{\sum(v)^2}{(Nl)^2}.
%$$
%\todo{someone should define these variables}

The number of free parameters in this model is large: we must optimize every grid point in the mean spectral templates and telluric basis vectors along with the stellar \RV s and the telluric basis weights. 
While in principle degeneracies between the stellar and telluric spectra should not exist since the star will always vary in \RV by at least \BERV \todo{(uhhh is this true??)}, in practice the complexity of the optimization problem makes parameter degeneracy a significant concern \todo{(total BS)}. 
We use a combination of L1 and L2 regularization to mitigate this issue. 
\todo{define these, discuss their purpose}

The final likelihood to be optimized is then:
\begin{equation}
\begin{split}
\ln \mathcal{L} = & \sum_{n} \sum_{m} -0.5 (y_{m,n} - y'_{m,n})^T C^{-1} (y_{m,n}-y'_{m,n})  \\
 & + \alpha_{L1, \mu_{\star}}  \sum_{m'} | \mu_{{\star}} | + \alpha_{L2, \mu_{\star}}  \sum_{m'} \mu_{{\star}}^2 \\
 & + \alpha_{L1, \mu_t}  \sum_{m'} | \mu_{t} |  + \alpha_{L2, \mu_t}  \sum_{m'} \mu_{t}^2 \\
 & + \alpha_{L1, W_t}  \sum_{k} \sum_{m'} | W_{t} |  + \alpha_{L2, W_t}  \sum_{k} \sum_{m'} W_{t}^2 + \alpha_{L2, z_t} \sum_{k} \sum_{n} z_{t}^2 ,
\end{split}
\end{equation}
where $\alpha_{L2, z_t}$ is set to 1.0 and all other regularization amplitudes are set by a validation procedure \todo{as described below}.

\subsection{Optimizing the Model}

We build an initial template for the stellar spectrum by setting $v_{obs} = v_{BERV}$ for all epochs. We can then calculate the median flux across \BERV-corrected spectra in bins at each model wavelength $\lambda'$. The telluric template is then initialized using the residuals after the stellar contribution has been removed, again by binning (this time without applying any Doppler shift to the spectra) and taking the median values of each bin. Finally, we initialize the telluric spectrum's basis vectors $W_t$ and weights $z_t$ by performing Principal Component Analysis (\acronym{PCA}) on the residuals after both the stellar spectrum and the mean telluric spectrum have been removed. The $K$ highest eigenweights and their corresponding eigenvectors are taken as the basis weights and vectors. \todo{(math here?)}

@@

The above-described model can be implemented in a variety of ways. We chose to build our code, \wobble, using the \TF package \todo{[terminology?]} in \code{python}. 
\TF is a \todo{[blah blah technical blah]}. 
While it has primarily been used in machine-learning contexts in the astronomical literature \todo{(cite)}, its auto-differentiation and fast linear algebra capabilities make it valuable for this application as well. 
The necessary optimizations over many parameters can be performed with high efficiency by \TF: the below-described analysis of \todo{x} spectra of 51 Peg runs in \todo{y} minutes on a standard Mac desktop.
